### 2020-08-05 ART_Piechowicz/Piechowicz-PU.tex
P. Polak correctons added

\begin{artengenv}{Robert Piechowicz}
	{A Formal Reconstruction of the Notions of Belief, Utterance and Trust}
	{A Formal Reconstruction of the Notions of Belief, Utterance and Trust}
	{A Formal Reconstruction of the Notions of Belief, Utterance\\and Trust}
	{Pontifical University of John Paul II in Krakow, Poland}
	{Problem of epistemic activity and their relationship with language is very well known in philosophy. Undertaking this challenge in this article we shall present some logical constructions apparent in these issues. More precisely we want to describe some difficulties of formal reconstruction of the notion of belief and utterance and try to find broader perspective appointed by notion of trust. To realize this goals article shows how non-formal assumption about \textit{doxa} affects on its  formal construction. Then, logic of utterances---mainly based on Conversational Implicatures theory---and its relation to doxastic systems is discussed and, finally, article shows that more accurate description of these two notions needs broader perspective created by BIT system proposed by Ch-J. Liau.}
	{logic, beliefs, utterances, trust.}

\indent The problem of the epistemic activity and its relationship with language is well known in philosophy. Moreover, it is one of the discussed topics in logic and although many formally correct solutions are proposed, their application to the natural language is more than problematic. Taking up this challenge in the following article we will present a logical construction of these three issues: belief, utterance and trust. The first chapter bears a \textit{quasi}--historical sense and allows to make the notion of belief more precise. In second chapter some problems of the doxastic logic systems will be discussed and the manner in which the non-formal assumption about \textit{doxa} affect its formal construction will be presented. Third chapter will concern the logic of utterances and its relation to the doxastic systems. Finally, we will try to show that more accurate description of those two notions needs broader perspective created by trust logic.

@@

The logical reflection on the epistemic and doxastic topics has been interrupted for many years. Philosophy evolved into other issues and, furthermore, thinkers did not have suitable tools for a more advanced analysis. The modern treatments of the logic of knowledge and belief arose after the Second World War and grew out of the work of such logicians as: Rudolf Carnap, Jerzy \L{}o\'{s}, Arthur Prior, Nicholas Rescher, G.H. von Wright. Von Wright suggested that any analysis of sentences about ``knowledge'' and ``belief'' needs the axiomatic approach. The logical systems developed by these thinkers were focused on ``relational operator $Kxp$ for ``$x$ knows $p$, where $Kx$ can be thought of as a parametrized modality characterizing  the person-relative epistemic status of a proposition''
%tu
%(Reschner, 2002, p , 478).
\parencite[][p.478]{jacquette_epistemic_2002}
The same kind of analysis was applied to belief and later to other epistemic states.

@@

The third feature---accessibility---is a stronger condition then disposition. ``This is a matter not of what a person would say if asked (= dispositional knowledge) but what one could say if he is sufficiently clever about using the information that is at one’s disposal occurrently or dispositionally''
%tu
(Reschner, 2002, p , 478).
\parencite[][p.478]{jacquette_epistemic_2002}
The logical analysis of cognitive states is founded on this feature because it refers not to a real subject which can err in information processing but to an idealized subject which can apply any logical schema to infer proper conclusion from the given evidence.

>>>>

\begin{artengenv}{Robert Piechowicz}
	{A formal reconstruction of the notions of belief, utterance and trust}
	{A formal reconstruction of the notions of belief, utterance and trust}
	{A formal reconstruction of the notions of belief, utterance\\and trust}
	{Pontifical University of John Paul II in Krakow, Poland}
	{Problem of epistemic activity and their relationship with language is very well known in philosophy. Undertaking this challenge in this article we shall present some logical constructions apparent in these issues. More precisely we want to describe some difficulties of formal reconstruction of the notion of belief and utterance and try to find broader perspective appointed by notion of trust. To realize this goals article shows how non-formal assumption about \textit{doxa} affects on its  formal construction. Then, logic of utterances---mainly based on Conversational Implicatures theory---and its relation to doxastic systems is discussed and, finally, article shows that more accurate description of these two notions needs broader perspective created by BIT system proposed by Ch-J. Liau.}
	{logic, beliefs, utterances, trust.}

\indent \lettrine[loversize=0.13,lines=2,lraise=-0.05,nindent=0em,findent=0.2pt]%
{T}{}he problem of the epistemic activity and its relationship with language is well known in philosophy. Moreover, it is one of the discussed topics in logic and although many formally correct solutions are proposed, their application to the natural language is more than problematic. Taking up this challenge in the following article we will present a logical construction of these three issues: belief, utterance and trust. The first chapter bears a \textit{quasi}--historical sense and allows to make the notion of belief more precise. In second chapter some problems of the doxastic logic systems will be discussed and the manner in which the non-formal assumption about \textit{doxa} affect its formal construction will be presented. Third chapter will concern the logic of utterances and its relation to the doxastic systems. Finally, we will try to show that more accurate description of those two notions needs broader perspective created by trust logic.

@@

The logical reflection on the epistemic and doxastic topics has been interrupted for many years. Philosophy evolved into other issues and, furthermore, thinkers did not have suitable tools for a more advanced analysis. The modern treatments of the logic of knowledge and belief arose after the Second World War and grew out of the work of such logicians as: Rudolf Carnap, Jerzy \L{}o\'{s}, Arthur Prior, Nicholas Rescher, G.H. von Wright. Von Wright suggested that any analysis of sentences about ``knowledge'' and ``belief'' needs the axiomatic approach. The logical systems developed by these thinkers were focused on ``relational operator $Kxp$ for ``$x$ knows $p$, where $Kx$ can be thought of as a parametrized modality characterizing  the person-relative epistemic status of a proposition''
%tu
%(Reschner, 2002, p , 478).
\parencite[][p.478]{jacquette_epistemic_2002}.
The same kind of analysis was applied to belief and later to other epistemic states.

@@

The third feature---accessibility---is a stronger condition then disposition. ``This is a matter not of what a person would say if asked (= dispositional knowledge) but what one could say if he is sufficiently clever about using the information that is at one’s disposal occurrently or dispositionally''
%tu
%(Reschner, 2002, p , 478).
\parencite[][p.478]{jacquette_epistemic_2002}.
The logical analysis of cognitive states is founded on this feature because it refers not to a real subject which can err in information processing but to an idealized subject which can apply any logical schema to infer proper conclusion from the given evidence.

### 2020-08-05 ART_Szybka/Szybka-PU.tex
P. Polak correctons added

\section{Introduction}

The mysterious relation between mathematics and physical world was and still is a subject of a debate \parencite{wigner_unreasonable_1960}. From the standpoint of a theoretical physicist, but not a professional philosopher, I must admit that existence of black holes is one of the sharpest proofs of the idea that underlying structure of the physical world is of the mathematical nature.

@@

The scientific impact of the first image of a black hole may be less profound than the impact of the recent gravitational waves detections, but seeing is believing. The image of the black hole presented to the public by the Event Horizon Telescope Collaboration\footnote{The results were published on the 10th of April in the six papers in a special issue of \textit{The Astrophysical Journal Letters}, \parencite[see e.g.][]{the_event_horizon_telescope_collaboration_first_2019}. More about Event Horizon Telescope Collaboration, \parencite[see][]{noauthor_event_nodate}.} is groundbreaking for the perception of black holes as real objects, especially beyond the scientific community.

@@

As we known today, astrophysical black holes exist in two sizes. The so-called stellar black holes with masses ranging from several to dozens of solar masses are created as an end state of evolution of massive stars. The so-called galactic black holes (or supermassive black holes) with masses ranging from hundreds of thousands to tens of billions of solar masses reside in the center of galaxies. The black holes are black, so good candidates for taking an image must be active---they must swallow the surrounding matter:  a cosmic dust, a gas, stars. The matter falling on a black hole radiates huge amount of energy which lighten up the neighbourhood and creates the black hole shadow. The black hole shadow contains the event horizon, thus we expected to see a gate to the region of `no return.' The candidates must also have possibly large angular size---this parameter is crucial for resolving images. It turns out that in this competition the supermassive black holes are clear winners. The Event Horizon Telescope team picked out two candidates: the nearest supermassive black hole Sgr A* at the center of our galaxy and much bigger supermassive black hole at the center of the galaxy M87. Our supermassive black gole Sgr A* is about 26000 light years away from Earth and has about four million solar masses. The M87 black hole is about 1500 times more massive and 2000 times further away. It follows from this comparison that the size of the M87 black hole event horizon is only about 22 micro-arc-sec of a degree compared to 53 micro-arc-sec of a degree of Sgr A*. Nevertheless, the M87 black hole turned out to be a better candidate. Although the Event Horizon Telescope team attempted to create images of both black holes only the image of the M87 black hole was presented on the 10th of April 2019. The center of our galaxy is not a best direction to conduct this kind of imaging. The larger mass of the M87 black hole implies that the image change less during observation time which guarantees more stable condition for taking an image. Photographing a running athlete with long exposure time is definitely much harder than taking a photo of grandparents behind a table!

@@

The image obtained by the Event Horizon Telescope Collaboration (with 200 scientist involved) was a spectacular example of application of very large base interferometry. The breakthrough in technology and new algorithms allowed the team to achieve something that not so long time ago was presumed to be impossible. What is even more important: the future is bright. The first image of the black hole which was presented on the 10th of April 2019 was not a final product of the project, but just a first outcome. With more instruments joining the Event Horizon Telescope Collaboration and increased sensitivity we should expect more spectacular images of the M87 black hole and other black holes. Probably, in not so distant future, we may even expect  movies showing the black holes in action.

>>>>

\section{Introduction}
\indent \lettrine[loversize=0.13,lines=2,lraise=-0.05,nindent=0em,findent=0.2pt]%
{T}{}he mysterious relation between mathematics and physical world was and still is a subject of a debate \parencite{wigner_unreasonable_1960}. From the standpoint of a theoretical physicist, but not a professional philosopher, I must admit that existence of black holes is one of the sharpest proofs of the idea that underlying structure of the physical world is of the mathematical nature.

@@

The scientific impact of the first image of a black hole may be less profound than the impact of the recent gravitational waves detections, but seeing is believing. The image of the black hole presented to the public by the Event Horizon Telescope Collaboration\footnote{The results were published on the 10\textsuperscript{th} of April in the six papers in a special issue of \textit{The Astrophysical Journal Letters}, \parencite[see e.g.][]{the_event_horizon_telescope_collaboration_first_2019}. More about Event Horizon Telescope Collaboration, \parencite[see][]{noauthor_event_nodate}.} is groundbreaking for the perception of black holes as real objects, especially beyond the scientific community.

@@

As we known today, astrophysical black holes exist in two sizes. The so-called stellar black holes with masses ranging from several to dozens of solar masses are created as an end state of evolution of massive stars. The so-called galactic black holes (or supermassive black holes) with masses ranging from hundreds of thousands to tens of billions of solar masses reside in the center of galaxies. The black holes are black, so good candidates for taking an image must be active---they must swallow the surrounding matter:  a cosmic dust, a gas, stars. The matter falling on a black hole radiates huge amount of energy which lighten up the neighbourhood and creates the black hole shadow. The black hole shadow contains the event horizon, thus we expected to see a gate to the region of `no return.' The candidates must also have possibly large angular size---this parameter is crucial for resolving images. It turns out that in this competition the supermassive black holes are clear winners. The Event Horizon Telescope team picked out two candidates: the nearest supermassive black hole Sgr A* at the center of our galaxy and much bigger supermassive black hole at the center of the galaxy M87. Our supermassive black gole Sgr A* is about 26000 light years away from Earth and has about four million solar masses. The M87 black hole is about 1500 times more massive and 2000 times further away. It follows from this comparison that the size of the M87 black hole event horizon is only about 22 micro-arc-sec of a degree compared to 53 micro-arc-sec of a degree of Sgr A*. Nevertheless, the M87 black hole turned out to be a better candidate. Although the Event Horizon Telescope team attempted to create images of both black holes only the image of the M87 black hole was presented on the 10\textsuperscript{th} of April 2019. The center of our galaxy is not a best direction to conduct this kind of imaging. The larger mass of the M87 black hole implies that the image change less during observation time which guarantees more stable condition for taking an image. Photographing a running athlete with long exposure time is definitely much harder than taking a photo of grandparents behind a table!

@@

The image obtained by the Event Horizon Telescope Collaboration (with 200 scientist involved) was a spectacular example of application of very large base interferometry. The breakthrough in technology and new algorithms allowed the team to achieve something that not so long time ago was presumed to be impossible. What is even more important: the future is bright. The first image of the black hole which was presented on the 10\textsuperscript{th} of April 2019 was not a final product of the project, but just a first outcome. With more instruments joining the Event Horizon Telescope Collaboration and increased sensitivity we should expect more spectacular images of the M87 black hole and other black holes. Probably, in not so distant future, we may even expect  movies showing the black holes in action.
